---
editor_options: 
  markdown: 
    wrap: 72
---

# Superstore EDA

```{r}
# Setup
library(tidyr)
library(tidyverse)
library(ggplot2)

superstore <- read.csv("Sample - Superstore.csv")
```

```{r}
# Exploratory analysis
str(superstore)
```

```{r}
#Checking for null values
sum(is.na(superstore))
# Checking for duplicates
sum(duplicated(superstore))

# convert columns to correct types
superstore$Order.Date <- as.Date(superstore$Order.Date, format = "%m/%d/%Y")
superstore$Ship.Date <- as.Date(superstore$Ship.Date, format = "%m/%d/%Y")

# Checking for data consistency
table(superstore$Category)
table(superstore$Region)
table(superstore$Ship.Mode)

```

```{r}
#creating new month and year column
superstore <- superstore %>%
  mutate(Year = format(`Order.Date`, "%Y")) %>%
  mutate(Month = format(`Order.Date`, "%m")) %>%
  mutate(YearMonth = format(`Order.Date`, "%Y-%m"))

superstore$Year <- as.numeric(as.character(superstore$Year))  # Keep Year numeric
```

```{r}
# handling outliers
superstore_numeric <- superstore %>%
  select(Order.ID, Sales, Profit)

superstore_numeric_long <- superstore_numeric %>%
  pivot_longer(cols = c(Sales, Profit), 
               names_to = "Variable",
               values_to = "Values")

outliers_box <- ggplot(superstore_numeric_long, aes(Values, Variable))
outliers_box + geom_boxplot()
```

```{r}
# checking outliers
sales_outlier <- superstore %>%
  subset(Sales > 5000)
```

-\> All based on high quantity, technology purchases and hence not
removed

```{r}
# checking outliers
profit_outlier <- superstore %>%
  subset(Profit > 3000 | Profit < -2500)
```

-\> Negative outliers because of large discounts and positive outliers
align with sales outliers

# Question 1: How does sales performance vary over time?

## Objective: Identify monthly trends and seasonal patterns.

### Tasks:

```{r}
# group sales by month and year
sales_by_month_year <- superstore %>%
  group_by(YearMonth) %>%
  summarize(TotalSales = sum(Sales, na.rm = TRUE))

# create a table with sales by year and month
sales_by_month_year_split <- superstore %>%
  select(Sales, Month, Year)
sales_by_month_year_split_wide <- sales_by_month_year_split %>%
  pivot_wider(names_from = Month, values_from = Sales, 
              values_fn = sum, names_prefix = "Month_")

sales_by_month_year_split_wide <- sales_by_month_year_split_wide %>%
  select(Year, sort(names(.)[-1]))

# Reshape data back to long format for plotting
sales_by_month_year_split_long <- sales_by_month_year_split_wide %>%
  pivot_longer(cols = starts_with("Month_"), names_to = "Month", 
               values_to = "Sales") %>%
  mutate(Month = as.integer(gsub("Month_", "", Month)))
```

-   Identify which months consistently show sales peaks

```{r}
# Plotting
ggplot(sales_by_month_year, aes(x = YearMonth, y = TotalSales)) +
  geom_col() +
  labs(title = "Sales Over Time", x = "Month-Year", y = "Total Sales") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

-\> Shows sales peaks towards end of the year

```{r}
# normalise for better compariseon
sales_by_month_year_split_long <- sales_by_month_year_split_long %>%
  group_by(Year) %>%
  mutate(Sales_Scaled = (Sales - min(Sales)) / (max(Sales) - min(Sales))) %>%
  ungroup()
# output in heatmap
heatmap_plot <- ggplot(sales_by_month_year_split_long, aes(x = factor(Month), y = factor(Year), fill = Sales_Scaled)) 
heatmap_plot +  geom_tile() +
  scale_fill_viridis(name = "Sales", option = "C") +
  labs(x = "Month", y = "Year", title = "Sales Heatmap") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

-\> Normalised sales per month & year confirm peaks for last 4 months of
the year with November and December dominating. Surprising small peak as
well in March.

```{r}
# multiple regression with year as continuous and month as dummy variable
sales_by_month_year_split_long$Month <- as.factor(sales_by_month_year_split_long$Month)  # Ensure Month is categorical
sales_by_month_year_split_long$Year <- sales_by_month_year_split_long$Year - min(sales_by_month_year_split_long$Year)  # Shift year so the first year is 0
model_final <- lm(Sales ~ Year + Month, data = sales_by_month_year_split_long)
summary(model_final)
```

-\> Multiple regression shows significant increase in sales per year and
confirms the significance of the large peaks at the end of the year
(September, November and December) and also shows significant deviation
in March.

```{r}
ggplot(sales_by_month_year_split_long, aes(x = Year, y = Sales)) +
  geom_point() +  # Scatter plot of sales data
  geom_smooth(method = "lm", se = TRUE, color = "blue") +  # Regression line
  labs(title = "Sales Over Years",
       x = "Years (Starting at 0)",
       y = "Sales") +
  theme_minimal()

ggplot(sales_by_month_year_split_long, aes(x = as.factor(Month), y = Sales)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Monthly Sales Seasonality",
       x = "Month",
       y = "Sales") +
  theme_minimal()
```

-   Calculate YOY growth rate for the most recent complete year

```{r}
# grouping by year
sales_by_year <- superstore %>%
  select(Sales, Year) %>%
  group_by(Year) %>%
  summarise(TotalSales =sum(Sales))

#calculating growth
sales_by_year <- sales_by_year %>%
  mutate(YoY_Growth = round((TotalSales / lag(TotalSales) -1)*100,2)) %>%
  subset(Year > 2014)

# plotting
growth_plot <- ggplot(sales_by_year, aes(x=factor(Year), y=YoY_Growth, fill=YoY_Growth > 0)) 
growth_plot+
  geom_col() +
  scale_fill_manual(values=c("FALSE"="red", "TRUE"="darkgreen")) +
  labs(title = "YoY Growth Rate", y="Growth in %", x="Year") +
  geom_text(aes(label=sprintf("%.1f", YoY_Growth))) +
  theme_minimal() + theme(legend.position = "none")
```

-   Suggest potential reasons for seasonal patterns

    --\> Potential reasons include christmas for November and December
    and end of season sales in winter (March) and summer (September)

```         
```

# Question 2: Which product categories have the best/worst profit margins?

## Objective: Identify high/low performing product categories

### Tasks:

-   Identify which category has the thinnest margins

```{r}
# create df for categories
profit_margin_by_cat <- superstore %>%
  select(Category, Sales, Profit) %>%
  group_by(Category) %>%
  summarise(TotalSales = sum(Sales), TotalProfit =sum(Profit)) %>%
  mutate(ProfitMargin = round(TotalProfit/TotalSales*100,2)) %>%
  arrange(ProfitMargin)

cat_lowest_margin = slice(profit_margin_by_cat,1)
cat_lowest_margin
```

-\> The Furniture Category has the lowest Margins with 2.49%

-   Calculate the profit margin difference between top and bottom
    categories

```{r}
highest_margin = profit_margin_by_cat[nrow(profit_margin_by_cat),]$ProfitMargin
highest_margin - cat_lowest_margin$ProfitMargin
```

-\> The difference is 14.91% between the Furniture(lowest) and
Technology (highest) Category Profit Margin

-   Suggest strategies to improve low-performing categories

-\> Check for subcategories within the category to see if there are
products that can be dropped, check price increases or cost reduction
measurements. At the same time, the sales volume for the category must
be checked to make any early conclusions. It could be that the category
has much higher sales rates and naturally lower profit margins but still
resulting in high overall Profit.

# Question 3: How do regional performances compare?

## Objective: Compare sales distribution and profitability across regions

### Tasks:

-   Identify which region has both high sales and high profitability

```{r}
# create df for categories
performance_by_region <- superstore %>%
  select(Region, Sales, Profit) %>%
  group_by(Region) %>%
  summarise(TotalSales = sum(Sales), TotalProfit =sum(Profit)) %>%
  mutate(ProfitNormalised = TotalProfit / sum(TotalProfit),
         SalesNormalised = TotalSales / sum(TotalSales), 
         Performance = ProfitNormalised + SalesNormalised) %>%
  mutate(ProfitMargin = round(TotalProfit/TotalSales*100,2)) %>%
  arrange(desc(Performance))

performance_by_region[1,]$Region

```

The Region "West" shows the highest Sales and Profitability

-   Find any regions with negative profits

```{r}
regions_with_loss <- performance_by_region %>%
  subset(TotalProfit < 0)
regions_with_loss

```

-\> There are no regions with negative profits

-   Analyze if high sales always correlate with high profits

```{r}
sale_profit_plot <- ggplot(performance_by_region, aes(x=TotalSales, y=TotalProfit))
sale_profit_plot + geom_point() + geom_smooth(method = "lm")
```

```{r}
# Fit the linear regression model
model <- lm(TotalSales ~ TotalProfit, data = performance_by_region)

# Display the summary of the model
summary(model)
```

-\> Correlation of 0.85 shows strong relation between Sales and Profit.
However, with a p-value of 7%, this is statistically insignificant

-   Propose regional-specific strategies based on findings

-\> Profit Margin of the Central Region seems to be the only clear
outlier and needs to be further evaluated. Regional performance needs to
be contrasted to factors such as market size and market share in the
various regions to derive appropriate conclusions on the differing
values in the analysis.

# Question 4: What does customer segmentation reveal?

## Objective: Identify valuable customer groups using RFM analysis

### Tasks:

-   Calculate percentage of customers in each segment
-   Identify which segment generates the most revenue

```{r}
customer_segment <- superstore %>%
  select(Customer.ID, Segment, Sales) %>%
  group_by(Segment) %>%
  summarise(Customers = n_distinct(Customer.ID), TotalSales = sum(Sales)) %>%
  mutate(CustomersShare = round(Customers/ sum(Customers)*100,2))
print(customer_segment)
```

-   Develop retention strategies for “At Risk” customers

```{r}
library(dplyr)
library(lubridate)

# Get relevant customer data
at_risk <- superstore %>%
  select(Order.Date, Customer.ID, Sales) %>%
  group_by(Customer.ID) %>%
  summarise(
    LastPurchase = max(Order.Date),
    Spending = sum(Sales),
    TotalPurchases = n(),
    NumberPurchasesLast3Months = sum(Order.Date >= as.Date("2018-01-01") - months(3)),
    NumberPurchases3_6Months = sum(Order.Date >= as.Date("2017-10-01") - months(3)),
    PurchaseHistoryDiff = NumberPurchasesLast3Months - NumberPurchases3_6Months,
    SalesLast3Months = sum(Sales[Order.Date >= as.Date("2018-01-01") - months(3)]),
    Sales3_6Months = sum(Sales[Order.Date >= as.Date("2017-10-01") - months(3)]),
    SalesHistoryDiff = SalesLast3Months - Sales3_6Months
  )

# Add recency and scores
at_risk <- at_risk %>%
  mutate(
    Recency = as.numeric(as.Date("2018-01-01") - LastPurchase),
    RecencyScore = case_when(
      Recency <= 30 ~ 3,
      Recency <= 60 ~ 2,
      Recency <= 90 ~ 1,
      Recency > 90 ~ 0
    ),
    FrequencyScore = case_when(
      NumberPurchasesLast3Months != 0 & (NumberPurchasesLast3Months - PurchaseHistoryDiff > 2) ~ 3,
      NumberPurchasesLast3Months != 0 & (NumberPurchasesLast3Months - PurchaseHistoryDiff > 0) ~ 2,
      NumberPurchasesLast3Months != 0 & (NumberPurchasesLast3Months - PurchaseHistoryDiff > -2) ~ 1,
      TRUE ~ 0
    ),
    SalesScore = case_when(
      SalesLast3Months != 0 & (SalesLast3Months - SalesHistoryDiff > 20) ~ 3,
      SalesLast3Months != 0 & (SalesLast3Months - SalesHistoryDiff > 0) ~ 2,
      SalesLast3Months != 0 & (SalesLast3Months - SalesHistoryDiff > -20) ~ 1,
      TRUE ~ 0
    )
  )

```

-   Suggest marketing approaches for “High Spenders”

```{r}
summary(at_risk$SalesHistoryDiff)

```

# Question 5: How does shipping mode affect profitability?

## Objective: Analyze cost-to-serve across shipping modes

### Tasks:

-   Compare profit margins across shipping modes

```{r}

```

-   Calculate profit per order for each shipping mode

```{r}

```

-   Suggest optimal shipping strategy based on findings

```{r}

```

## Extra Challenge:

-   Identify 3 actionable business recommendations.

```{r}

```

-   Propose 2 new questions for deeper analysis.

```{r}

```

-   Estimate potential financial impact of your recommendations.

```{r}

```

## Discussion Points:

-   How do sales trends correlate with marketing initiatives?
-   Are there regional preferences for product categories?
-   What operational changes could improve low-margin categories?
-   How might customer segmentation affect inventory management?
